Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-04-08 paper.tex
pop studies finished

\section{Predictions and observational limits}

@@

@@

@@

@@

To complete this section

if CC, then skewed towards large X-ray fluxes; X-ray emission good
proxies for gamma-ray emission --) The dust turns up at about a factor
100 lower than the stars --) The other figure show the spectral shape
of the IC-star with the same spectral shape as the dust, it shows the
same gamma-ray slope as the dust --) The reason why there is only a
factor 10 difference in the break in the gamma-ray emission between
the star and dust is related to the KN. In our CRp paper, the last
figure shows that at $\eg=100\,\tev$, the spectra is suppressed with
about a factor 10 (also depends on $\alpha_\e$). Since the dust
photons have a factor 10 higher energy, I expect this suppression at
about 10~TeV. For star photons the suppression should be at
100~GeV. So the star photons are definitely KN suppressed, which
explains the earlier break and the steeper slope.

@@

In this paper, we study the possibility for detecting $\gamma$-ray emission in
galaxy clusters. We consider benchmark and leptophilic models of supersymmetric
dark matter as well as cosmic ray induced pion decay which is thought to
dominate the astrophysical signal from clusters. Once supersymmetric cold dark
matter decouples from the expanding universe it streams freely and smooths over
any potentially present smaller scales. This imprints as a characteristic
minimum halo mass $M_\mathrm{min}$ into the hierarchy of structure formation and
should still survive until the present time as substructure within larger
halos. It turns out that the boost due to substructures dominates over the
smooth component for halo masses $M>10^3 \msun$ and contributes a constant
luminosity for each decade in halo mass. The ratio of halo-to-minimum subhalo
mass $M_{200}/M_\mathrm{min}$ is maximized for the largest systems that had time
to collapse and virialize since the Big Bang -- galaxy clusters. The
corresponding substructure boost can reach values exceeding $10^3$ which should
make nearby clusters the brightest DM annihilation sources after the Milky May
center and hypothetical very close-by DM substructures. Since the substructures
are distributed according to an Einasto profile, the contribution to the
annihilation emission from substructures peaks around the virial radius. This
implies an almost flat surface brightness profile of the annihilation emission
which makes it necessary to have the entire cluster in the field-of-view to take
advantage of the entire substructure boost.  This property makes it difficult to
detect DM annihilation emission without an enhancement of the particle physics
cross section over its standard value of $\sigma v\sim 3\times 10^{-26}
~\mathrm{cm}^3~\mathrm{s}^{-1}$ with imaging air Cherenkov telescopes due to the
spatially extended boost from substructures of angular extend $\sim
0.5-1\deg$. In contrast, the sensitivity of Cherenkov telescopes drops linearly
with radius outside the point spread function $\sim0.1\deg$.

>>>>

\section{Population studies: flux predictions and observational limits}

@@

\subsection{Scaling relations}

@@

\subsection{DM annihilation}

@@

\subsection{CR-induced emission}

@@

To complete this section, we present our semi-analytical model predictions for
the CR-induced $\gamma$-ray flux for all clusters in the extended HIFLUGCS
sample in Fig.~\ref{fig19}. In the lower panel of this figure, we show the
relative difference between the $\gamma$-ray flux predicted by the
mass-luminosity scaling relation and to the flux computed using the
semi-analytical model. Using our semi-analytic modelling results in a scatter of
the expected $\gamma$-ray flux at a given mass. Cool core clusters have a denser
core which increases the target density of gas protons for the hadronic reaction
which in turn biases the expected $\gamma$-ray fluxes systematically higher. We
confirm this effect as all our outliers with increased fux in our semi-analytic
model relative to the scaling relation expectation are in fact cool core
clusters suggesting that teh X-ray emission should be a good proxy for the
expected $\gamma$-ray emission. Note however that there could be the
counter-acting effect of CR streaming in relaxed clusters which would tend to
decrease the $\gamma$-ray luminosity \cite{2011A&A...527A..99E}. Future careful
modelling is needed to quatify this effect.

\del{
ANDERS: do you need this?
--) The dust turns up at about a factor 100 lower than the stars
--) The other figure show the spectral shape of the IC-star with the same
spectral shape as the dust, it shows the same gamma-ray slope as the dust
--) The reason why there is only a factor 10 difference in the break in the
gamma-ray emission between the star and dust is related to the KN. In our CRp
paper, the last figure shows that at $\eg=100\,\tev$, the spectra is suppressed
with about a factor 10 (also depends on $\alpha_\e$). Since the dust photons
have a factor 10 higher energy, I expect this suppression at about 10~TeV. For
star photons the suppression should be at 100~GeV. So the star photons are
definitely KN suppressed, which explains the earlier break and the steeper
slope.
}

@@

In this paper, we study the possibility for detecting $\gamma$-ray emission in
galaxy clusters. We consider benchmark and leptophilic models of supersymmetric
dark matter as well as cosmic ray induced pion decay which is thought to
dominate the astrophysical signal from clusters. Once supersymmetric cold dark
matter decouples from the expanding universe it streams and erases any
potentially present smaller scales. Once gravitational instabilities causes
halos to collapse, this free-streaming scale imprints as a characteristic
minimum halo mass $M_\mathrm{min}$ into the hierarchy of structure formation and
should still survive until the present time as substructure within larger
halos. It turns out that the boost due to substructures dominates over the
smooth component for halo masses $M>10^3 \msun$ [Anders?] and contributes a constant
luminosity for each decade in halo mass. The ratio of halo-to-minimum subhalo
mass $M_{200}/M_\mathrm{min}$ is maximized for the largest systems that had time
to collapse and virialize since the Big Bang -- galaxy clusters. The
corresponding substructure boost can reach values exceeding $10^3$ which should
make nearby clusters the brightest DM annihilation sources after the Milky May
center and hypothetical very close-by DM substructures. Since the substructures
are distributed according to an Einasto profile, the contribution to the
annihilation emission from substructures peaks around the virial radius. This
implies an almost flat surface brightness profile of the annihilation emission
which makes it necessary to have the entire cluster in the field-of-view to take
advantage of the entire substructure boost.  This property makes it difficult to
detect DM annihilation emission without an enhancement of the particle physics
cross section over its standard value of $\sigma v\sim 3\times 10^{-26}
~\mathrm{cm}^3~\mathrm{s}^{-1}$ with imaging air Cherenkov telescopes due to the
spatially extended boost from substructures of angular extend $\sim
0.5-1\deg$. In contrast, the sensitivity of Cherenkov telescopes drops linearly
with radius outside the point spread function $\sim0.1\deg$.
